---
title: "Intro to Bayes"
author: "Robin Elahi"
date: "January 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Acknowledgements and sources

This tutorial is based on other tutorials that I found on the web. I pirated shamelessly from these awesome sources:

  - https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started
  - http://mc-stan.org/bayesplot/articles/index.html
  - https://github.com/rmcelreath/rethinking
  - https://github.com/DanOvando/learn-stan
  - https://github.com/danilofreire/r-scripts/blob/master/stan-logistic-regression.R

There are many texts on Bayesian data analysis. Here are two that I found quite helpful in getting started:

  - ["Bayesian Models: A Statistical Primer for Ecologists"](https://press.princeton.edu/titles/10523.html)
  - ["Statistical Rethinking"](http://xcelab.net/rm/statistical-rethinking/)

## Installing stan and rstan

We are going to use ``rstan`` to fit our Bayesian models. So, you'll need to install ``rstan`` before class. Go to ``http://mc-stan.org`` and follow the instructions for your platform. The biggest challenge is getting a C++ compiler configured to work with your installation of R. The instructions at the [rstan getting started website](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) are quite thorough. 

Installing Stan can be a bit complicated on some machines, especially if you haven't run programs that require a C++ compiler before (if that sentance doesn't make sense to you, this means you), so **PLEASE DON'T WAIT UNTIL THE MORNING OF THE WORKSHOP TO TRY AND INSTALL STAN**. Try to install it a few days beforehand, and if you can't get it to work let me know and I'll try to help. 

To check that it works, once you've succesfully done all the other steps please open R and run the code below. If it works, you're good to go!

```{r, echo = TRUE, eval = FALSE}
library(rstan)
stanmodelcode <- "
data {
  int<lower=0> N;
  real y[N];
} 

parameters {
  real mu;
} 

model {
  target += normal_lpdf(mu | 0, 10);
  target += normal_lpdf(y  | mu, 1);
} 
"

y <- rnorm(20) 
dat <- list(N = 20, y = y); 
fit <- stan(model_code = stanmodelcode, model_name = "example", 
            data = dat, iter = 2012, chains = 3, sample_file = 'norm.csv',
            verbose = TRUE) 
print(fit)
traceplot(fit)

# extract samples 
e <- extract(fit, permuted = TRUE) # return a list of arrays 
mu <- e$mu 

m <- extract(fit, permuted = FALSE, inc_warmup = FALSE) # return an array 
print(dimnames(m))

# using as.array directly on stanfit objects 
m2 <- as.array(fit)
```

### Other prep

You can access the data we use in this tutorial here:

https://raw.githubusercontent.com/elahi/coding/master/bayes_180215/interview_logistic_subset.csv

Create a new folder called ``bayes_180215``, drop this data file in it, and set your [working directory](http://rfunction.com/archives/1001) to this folder. 

You should also have the following R packages installed: ``lme4, ggplot2, dplyr, purrr, tidyr, broom, bayesplot``. 

I recommend RStudio - see [here](https://www.rstudio.com/products/rstudio/download/) for installation. 

Make sure that your version of R is up to date, and all your packages are up to date. 

Ready to roll. 
